Retrofitting Effect Handlers onto OCaml

### 2020-11-10 retro-concurrency.tex
updates

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
effect handlers to OCaml is to support (1) highly concurrent applications such
as web servers to be expressed naturally in \emph{direct-style} as opposed to
using callbacks and (2) efficient nested parallel programs in the vein of
Cilk~\cite{Cilk}, OpenMP~\cite{}, Habenero Java~\cite{Imam14}, X10~\cite{},
Intel TBB~\cite{IntelTBB}, parallel Haskell variants~\cite{} and parallel ML
variants~\cite{}.

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
is meaningless and hence, we would lose compatibility with tools that inspect
the program stack. Hence, we choose not to use CPS translation and represent
continuations as call stacks.

@@

Our work is also timely. WebAssembly~\ref{} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables using an automated validator tool~\cite{}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
  \item extensive evaluation which shows that our implementation has negligible
    impact on code that does not use effect handlers, and serves as an
    efficient foundation for scalable concurrent and parallel programming.
\end{itemize}

@@

In this section, we refine the effect handler design for the OCaml use case.
The primary motivation for adding effect handlers to OCaml is to support
non-local control flow mechanism to implement asynchronous I/O in direct style,
generators and enable task parallelism. These use cases do not resume the
continuations more than once. Hence, our continuations are one-shot, and
resuming the continuation more than once raises |Invalid_argument| exception.
It is well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96RepresentingControl}.

While OCaml permits throwing exceptions across C frames, we do not allow
effects to propagate across C frames as the C frames would become part of the
captured continuation. Managing C frames as part of the continuation is a
complex endeavour~\cite{Daan17Aplas}, and we find that the complexity budget
outweighs the relatively fewer mechanisms enabled by this addition in our
setting.

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect
safety~\cite{Links,Koka,Effekt,Frank,HeliumBiernacki}. This is important for
backwards compatibility as our goal is to retrofit effect handlers to a
language with large legacy codebases. Programs that do not use effects remain
well-typed, and those that do compose well with those that don't.

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{} |'a eff| of effect values,
which when performed using the |perform| primitive returns a |'a| value.
Constructors for the value of type |'a eff| are declared using the effect
declarations. For example, the declaration |effect E : string -> int| is
syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be |continue|d with a
suitably typed value or |discontinue|d with an exception.

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen} for effect handlers, following the abstract machine
semantics of Hillerstrom et al.~\cite{Hillerstrom2020}. The crucial difference
from Hillerstrom et al. is that our stacks are composed of alternating sequence
of OCaml and C stack segments. The program state is captured as configuration
$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
under evaluation, its environment $\env$ and the current stack $\stack$. The
term is either an expression $e$ or a value $v$. The value are integer
constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect being
performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
environment is a map from variables to values.

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation. Observe
that the captured continuation |k''| includes the current handler. Intuitively,
the handler wraps around captured continuation. This gives Multicore OCaml
effect handlers deep handler semantics~\cite{}. \textsc{EffFwd} forwards the
effect to the outer fiber, and extends the captured continuation |k''| in the
process. Recall that we do not capture C frames as part of a continuation. To
this end, \textsc{EffUnHn} models unhandled effect. If the effect bubbles up to
the top fiber, which we know does not have an effect case (see
\textsc{Callback} rule), we raise \textsc{unhandled} exception at the point
where the corresponding effect was performed. This is achieved by appending the
captured continuation to the front of the current continuation.

@@

Unlike the operational semantics, handlers with just exception patterns
(exception handlers) are implemented differently from effect handlers. As
mentioned in \S~\ref{sec:exn_handlers}, exceptions are pervasive in OCaml and
are so cheap that they are used for local control flow. Hence, we retain the
linked exception handler frame implementation of stock OCaml exception handlers
in Multicore OCaml for performance backwards compatibility. This differs from
other research languages with effect handlers~\cite{Links,Frank,Eff}, which
implement exceptions using effects (by ignoring the continuation argument in
the handler).

@@

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
  with | v -> v | effect F kf -> ...
with | v -> v
		 | effect E ke ->
		     (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for Figure~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{2mm}
\mycaption{Layout of Multicore OCaml effect handlers.}
\end{figure*}

@@

The challenge with DWARF stack unwinding is to make it aware of the
non-contiguous stacks. While the complete details of DWARF stack unwinding is
beyond the scope of the paper, it is beneficial to know how DWARF unwind tables
are constructed in order to appreciate our solution. We refer the interested
reader to Bastian et al.~\cite{} for a good overview of DWARF stack unwinding.

@@

Our goal is to help DWARF identify the CFA when stacks are switched at external
calls, callbacks and effect handlers. We have implemented short, hand-written
DWARF bytecode at these three places, which we have validated with the DWARF
unwind table validator tool~\cite{Bastian}. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callback, we
save the current system stack pointer in the |DWARF block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented similarly by saving the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

@@

For stack walking during GC, the only change necessary was to make the runtime
aware of fiber layout and how to follow the parent pointers. The stack map
generation and scanning each frame for roots remains the same. Sivaramakrishnan
et al.~\cite{RetroParallel} present the challenges and solutions for
integrating fibers with the concurrent mark-and-sweep GC of Multicore OCaml.

@@

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. The
code does not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{} (|monad|) as a proxy for
CPS versions. Recall that OCaml compiler does not use CPS in its IR. In the
|monad| version, we use a |fork| to invoke the non-tail call and use an |MVar|
to collect its result.

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 13.77$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. Concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be GCed), where as effect handlers benefit from stack allocation of the frames.
Our concurrency monad, and indeed monadic concurrency libraries such as Lwt and
Async, also has other downsides; due to the lack of stack, exceptions,
backtraces, and DWARF unwinding are no longer useful.

We note that a compiler that uses CPS IR will be faster than the concurrency
monad implementation due to optimisations to reduce the heap allocation of
continuation frames. But Farvardin et al.~\cite{} show that CPS with
optimisations is still slower than using the call stack.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{48ns, 10ns, 13ns and 19ns}, respectively. For reference, the idle
memory latency for the local NUMA node is 93.2ns as measured using the Intel
MLC tool~\cite{}.

\paragraph{Generators} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{},
defunctionalised~\cite{DefunctionalizationAtWork} version (|cps|) and a
concurrency monad (|monad|) version of the generator for the tree. Both |cps|
and |monad| versions are specialised to the binary tree, and come with the
usual caveats of not using the stack for function calls. We observed that the
|cps| version was the fastest thanks to specialisation and hand optimisation.
|MC| version was only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

\paragraph{Chameneos} Chameneos~\cite{} is a concurrency game aimed at
measuring context switching and synchronization overheads. Our implementation
uses MVars~\cite{} for synchronization. We compare effect handler (|MC|),
concurrency monad (|monad|) and Lwt, a widely used concurrency programming
library for OCaml (|lwt|) versions. We observed that |MC| was the fastest, and
|monad| (|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.

\paragraph{Web server} As one of our primary goals of adding effect handlers to
OCaml is to support direct-style concurrency, we implement a full-fledged
HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
implement a user-level threading library, which spawns a thread per request. We
use \texttt{httpaf} library~\cite{} for HTTP handling, and
\texttt{libev}~\cite{} for the eventloop. We compare our implementation against
an Lwt version (|lwt|) which also uses \texttt{httpaf} and \texttt{libev}.
Unlike using effect handlers, Lwt version is written in monadic style, and does
not have a notion of thread per request. For comparison, we also include a Go
version (|go|) that uses \texttt{net/http}~\cite{} package.

@@

The client workload was generated with \texttt{wrk2}~\cite{}. We maintain 1k
open connections and perform requests for a static web page at different rates,
and record the request service rate and latency. The throughput and tail
latency graphs are given in Figures~\ref{grf:throughput} and~\ref{grf:latency}.
In all the versions, the throughput plateaus at around 30k requests per second.
We measure the tail latencies at 2/3rd of this rate (at 20k requests per
second), as would be the case in real deployments. We observe that both the
OCaml versions remain competitive with |go|, and |MC| performs best in terms of
tail latency.

Using effect handlers in the web server also helps with debugging the system as
a whole. Multicore OCaml supports getting the backtrace of a continuation,
which is a minor extension to getting the current backtrace. We can use this
feature to dump the backtrace of all the user-level threads in the web server.
Such a feature is provided by Go~\cite{}, but not possible with OCaml
concurrency libraries such as Async and Lwt which lack the notion of a thread.

@@

@@

\newpage

>>>>

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{everest,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{Madhavapeddy13}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
effect handlers to OCaml is to support (1) highly concurrent applications such
as web servers to be expressed naturally in \emph{direct-style} as opposed to
using callbacks and (2) efficient nested parallel programs in the vein of
Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP}, X10~\cite{Charles05}, Intel
TBB~\cite{IntelTBB}, and variants of Parallel ML~\cite{MaPLe, Fluet10,
Sivaramakrishnan14}.

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Farvardin20}. Moreover, with CPS, stack
unwinding is meaningless and hence, we would lose compatibility with tools that
inspect the program stack. Hence, we choose not to use CPS translation and
represent continuations as call stacks.

@@

Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{loom} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables using an automated validator tool~\cite{Bastian19}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
  \item extensive evaluation which shows that our implementation has negligible
    impact on code that does not use effect handlers, and serves as an
    efficient foundation for scalable concurrent and parallel programming.
\end{itemize}

@@

In this section, we refine the effect handler design for the OCaml use case.
The primary motivation for adding effect handlers to OCaml is to support
non-local control flow mechanism to implement asynchronous I/O in direct style,
generators and enable task parallelism. These use cases do not resume the
continuations more than once. Hence, our continuations are one-shot, and
resuming the continuation more than once raises |Invalid_argument| exception.
It is well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96}.

While OCaml permits throwing exceptions across C frames, we do not allow
effects to propagate across C frames as the C frames would become part of the
captured continuation. Managing C frames as part of the continuation is a
complex endeavour~\cite{Leijen17}, and we find that the complexity budget
outweighs the relatively fewer mechanisms enabled by this addition in our
setting.

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect
safety~\cite{Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is important for
backwards compatibility as our goal is to retrofit effect handlers to a
language with large legacy codebases. Programs that do not use effects remain
well-typed, and those that do compose well with those that don't.

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns a |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be |continue|d with a
suitably typed value or |discontinue|d with an exception.

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract machine
semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The crucial difference
from Hillerstrom et al. is that our stacks are composed of alternating sequence
of OCaml and C stack segments. The program state is captured as configuration
$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
under evaluation, its environment $\env$ and the current stack $\stack$. The
term is either an expression $e$ or a value $v$. The value are integer
constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect being
performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
environment is a map from variables to values.

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation. Observe
that the captured continuation |k''| includes the current handler. Intuitively,
the handler wraps around captured continuation. This gives Multicore OCaml
effect handlers deep handler semantics~\cite{Hillerstrom20}. \textsc{EffFwd} forwards the
effect to the outer fiber, and extends the captured continuation |k''| in the
process. Recall that we do not capture C frames as part of a continuation. To
this end, \textsc{EffUnHn} models unhandled effect. If the effect bubbles up to
the top fiber, which we know does not have an effect case (see
\textsc{Callback} rule), we raise \textsc{unhandled} exception at the point
where the corresponding effect was performed. This is achieved by appending the
captured continuation to the front of the current continuation.

@@

Unlike the operational semantics, handlers with just exception patterns
(exception handlers) are implemented differently from effect handlers. As
mentioned in \S~\ref{sec:exn_handlers}, exceptions are pervasive in OCaml and
are so cheap that they are used for local control flow. Hence, we retain the
linked exception handler frame implementation of stock OCaml exception handlers
in Multicore OCaml for performance backwards compatibility. This differs from
other research languages with effect handlers~\cite{Hillerstrom20,Frank,Eff}, which
implement exceptions using effects (by ignoring the continuation argument in
the handler).

@@

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
  with | v -> v | effect F kf -> ...
with | v -> v
		 | effect E ke ->
		     (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{2mm}
\mycaption{Layout of Multicore OCaml effect handlers.}
\end{figure*}

@@

The challenge with DWARF stack unwinding is to make it aware of the
non-contiguous stacks. While the complete details of DWARF stack unwinding is
beyond the scope of the paper, it is beneficial to know how DWARF unwind tables
are constructed in order to appreciate our solution. We refer the interested
reader to Bastian et al.~\cite{Bastian19} for a good overview of DWARF stack
unwinding.

@@

Our goal is to help DWARF identify the CFA when stacks are switched at external
calls, callbacks and effect handlers. We have implemented short, hand-written
DWARF bytecode at these three places, which we have validated with the DWARF
unwind table validator tool~\cite{Bastian19}. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callback, we
save the current system stack pointer in the |DWARF block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented similarly by saving the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

@@

For stack walking during GC, the only change necessary was to make the runtime
aware of fiber layout and how to follow the parent pointers. The stack map
generation and scanning each frame for roots remains the same. Sivaramakrishnan
et al.~\cite{Sivaramakrishnan20} present the challenges and solutions for
integrating fibers with the concurrent mark-and-sweep GC of Multicore OCaml.

@@

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. The
code does not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
proxy for CPS versions. Recall that OCaml compiler does not use CPS in its IR.
In the |monad| version, we use a |fork| to invoke the non-tail call and use an
|MVar| to collect its result.

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 13.77$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. Concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be GCed), where as effect handlers benefit from stack allocation of the frames.
Our concurrency monad, and indeed monadic concurrency libraries such as
Lwt~\cite{lwt} and Async~\cite{async}, also have other downsides; due to the
lack of stack, exceptions, backtraces, and DWARF unwinding are no longer
useful.

We note that a compiler that uses CPS IR will be faster than the concurrency
monad implementation due to optimisations to reduce the heap allocation of
continuation frames. But Farvardin et al.~\cite{Farvardin20} show that CPS with
optimisations is still slower than using the call stack.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{48ns, 10ns, 13ns and 19ns}, respectively. For reference, the idle
memory latency for the local NUMA node is 93.2ns as measured using the Intel
MLC tool~\cite{mlc}.

\paragraph{Generators} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a
concurrency monad (|monad|) version of the generator for the tree. Both |cps|
and |monad| versions are specialised to the binary tree, and come with the
usual caveats of not using the stack for function calls. We observed that the
|cps| version was the fastest thanks to specialisation and hand optimisation.
|MC| version was only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

\paragraph{Chameneos} Chameneos~\cite{Chameneos} is a concurrency game aimed at
measuring context switching and synchronization overheads. Our implementation
uses |MVars| for synchronization. We compare effect handler (|MC|), concurrency
monad (|monad|) and Lwt, a widely used concurrency programming library for
OCaml (|lwt|) versions. We observed that |MC| was the fastest, and |monad|
(|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.

\paragraph{Web server} As one of our primary goals of adding effect handlers to
OCaml is to support direct-style concurrency, we implement a full-fledged
HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
implement a user-level threading library, which spawns a thread per request. We
use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
style, and does not have a notion of thread per request. For comparison, we
also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
package.

@@

The client workload was generated with \texttt{wrk2}~\cite{wrk2}. We maintain
1k open connections and perform requests for a static web page at different
rates, and record the request service rate and latency. The throughput and tail
latency graphs are given in Figures~\ref{grf:throughput} and~\ref{grf:latency}.
In all the versions, the throughput plateaus at around 30k requests per second.
We measure the tail latencies at 2/3rd of this rate (20k requests per second)
to simulate optimal load. We observe that both of the OCaml versions remain
competitive with |go|, and |MC| performs best in terms of tail latency.

Using effect handlers in the web server also helps with debugging the system as
a whole. Multicore OCaml supports getting the backtrace of a continuation,
which is a minor extension to getting the current backtrace. We can use this
feature to dump the backtrace of all the user-level threads in the web server.
Such a feature is provided by Go~\cite{gopprof}, but not possible with OCaml
concurrency libraries such as Async and Lwt which lack the notion of a thread.

@@

\clearpage

@@

\clearpage
